RJObject Paper

### 2014-06-23 Documents/Paper/ms.tex
Embolden symbols

@@

\section{Introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Many inference problems have the following structure. There is some unknown
number, $N$, of objects in a region. Each of the $N$ objects has a property
$x$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a position and a mass).

We would like to assign a prior to the objects' properties $\{x_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``interim prior''
conditional on some hyperparameters $\alpha$, and then assign a prior to
$\alpha$. This kind of model is usually called {\it hierarchical}.

The prior for $N$, $\alpha$, and $\{x_i\}$ is usually factorised
in the following way:

\begin{eqnarray}
p(N, \alpha, \{x_i\}) &=& p(N) p(\alpha | N) p(\{x_i\} | \alpha, N) \\
&=& p(N) p(\alpha) \prod_{i=1}^N p(x_i | \alpha).
\end{eqnarray}

Here we have assumed the priors for $N$ and $\alpha$ are independent, and
the interim prior for $\{x_i\}$ is iid and does not depend on $N$.

@@

\section{Metropolis proposals for the general problem}
We will now define a set of proposal distributions that could be used to
sample the prior distribution
$p(N, \alpha, \{x_i\}) = p(N) p(\alpha) \prod_{i=1}^N p(x_i | \alpha)$.
These proposals will need to satisfy detailed balance in order to be valid
when used inside DNS. The proposals will also be very heavy tailed. When
DNS is exploring a distribution close to the prior, large proposals will
generally be needed. When DNS is exploring a distribution constrained to
very high values of the likelihood function, much smaller proposals will be
appropriate. Rather than attempting expensive tuning (since the number of
distributions involved may number in the hundreds or even thousands), we will
apply the heavy tailed proposal distributions and simply accept that there will
be some waste.

@@

\subsection{Proposals that modify $N$}\label{sec:proposal1}
The first kind of proposal we consider are proposals that change the
dimension of the model, i.e. proposals that change the value of $N$. By
necessity, these will also change the parameters $\{x_i\}_{i=1}^N$ because
the number of parameters will be changed. The proposals used here are
traditionally called {\it birth and death} proposals.

@@

If $N' > N$, the extra parameters needed, $\{x_i\}_{i=N+1}^{N'}$,
are drawn from their prior conditional on the current value of the
hyperparameters, i.e. the interim prior $p(x | \alpha)$.

@@

\subsection{Proposals that modify objects}\label{sec:proposal2}
We now consider a proposal distribution that modifies one or more of the
objects $\{x_i\}$, while keeping the number of objects $N$, as well as the
hyperparameters $\alpha$, fixed.

Let $F(x; \alpha)$ be a function that takes an object $x$ and transforms it
to a value $u$ that has a uniform distribution between 0 and 1, given $\alpha$.
If the object $x$ consists of a single scalar value, $F$ is the cumulative
distribution of the interim prior. Denote the inverse of $F$ by $G$.

A proposal
for an object involves transforming its parameters to $[0, 1]$ using $F$,
making the proposal in terms of $u$, and then transforming back.
Specifically, the proposal chooses
a new value $x_i'$ from the current value $x_i$ as follows:
\begin{eqnarray}
u_i &:=& F(x_i; \alpha)\\
u_i' &:=& \textnormal{mod}\left(u_i + \delta_u, 1\right)\\
x_i' &:=& G(u_i'; \alpha).
\end{eqnarray}
where $\delta_u$ is drawn from a heavy-tailed distribution.

@@

\subsection{Proposals that change the hyperparameters,
keeping the objects fixed}\label{sec:proposal3}
Another kind of proposal that we will consider is a proposal that keeps all of
the objects fixed in place (i.e. leaves $\{x_i\}$) unchanged, but changes
the hyperparameter(s) from their current value $\alpha$
to a new value $\alpha'$. The proposal for the hyperparameter(s) is chosen
to satisfy detailed balance with respect to $p(\alpha)$.

The overall Metropolis acceptance probability
for this kind of move, if sampling the prior (or the constrained prior of DNS)
must include the following factor:
\begin{eqnarray}
\frac{\prod_{i=1}^N p(x_i | \alpha')}{\prod_{i=1}^N p(x_i | \alpha)}
\label{eqn:acceptance_prob}
\end{eqnarray}
Since this proposal leaves the objects $\{x_i\}$ fixed, it will usually not
affect the value of the likelihood, and therefore the likelihood will not need
to be recomputed.

\subsection{Proposals that change the hyperparameters
and all of the objects}\label{sec:proposal4}
The above proposals allow for changes to $N$, $\alpha$, and $\{x_i\}$,
and are therefore sufficient to allow for ``correct'' exploration of the
prior distribution, and the constrained prior distributions used in DNS.
However, they do not necessarily allow for {\it efficient} exploration, even
of the prior itself. The main reason for this is the inability for large
changes to be made to the hyperparameters $\alpha$. If the proposed change
from $\alpha$ to $\alpha'$ is large, the ratio in
Equation~\ref{eqn:acceptance_prob} is likely to be very small, and the move
will probably be rejected.

Therefore, we allow an additional move that changes $\alpha$ to a new
value $\alpha'$, but rather than leaving the objects $\{x_i\}$ fixed, we
drag them so they represent the distribution $p(x | \alpha')$ rather than
$p(x|\alpha)$. We do this by making use of the transformation functions
$F(x; \alpha)$ and $G(x; \alpha)$ defined in Section~\ref{sec:proposal2}.

The ``dragging'' process works as follows, and must be carried out on
each object:
\begin{eqnarray}
u_i &:=& F(x_i; \alpha)\\
x_i' &:=& G(x_i; \alpha')
\end{eqnarray}

>>>>

\newcommand{\hyperparams}{\boldsymbol{\alpha}}
\newcommand{\xx}{\mathbf{x}}

@@

\section{Introduction}
\setlength{\parindent}{0cm}
\setlength{\parskip}{3mm}
Many inference problems have the following structure. There is some unknown
number, $N$, of objects in a region. Each of the $N$ objects has a property
$\xx$, which may be a single scalar value (for example, a mass), or a set of
values (e.g. a position and a mass).

We would like to assign a prior to the objects' properties $\{\xx_i\}_{i=1}^N$.
Since $N$ may be large, it is usually easier to assign an ``interim prior''
conditional on some hyperparameters $\hyperparams$, and then assign a prior to
$\hyperparams$. This kind of model is usually called {\it hierarchical}.

The prior for $N$, $\hyperparams$, and $\{\xx_i\}$ is usually factorised
in the following way:

\begin{eqnarray}
p(N, \hyperparams, \{\xx_i\}) &=& p(N) p(\hyperparams | N) p(\{\xx_i\} | \hyperparams, N) \\
&=& p(N) p(\hyperparams) \prod_{i=1}^N p(\xx_i | \hyperparams).
\end{eqnarray}

Here we have assumed the priors for $N$ and $\hyperparams$ are independent, and
the interim prior for $\{\xx_i\}$ is iid and does not depend on $N$.

@@

\section{Metropolis proposals for the general problem}
We will now define a set of proposal distributions that could be used to
sample the prior distribution
$p(N, \hyperparams, \{\xx_i\}) = p(N) p(\hyperparams) \prod_{i=1}^N p(\xx_i | \hyperparams)$.
These proposals will need to satisfy detailed balance in order to be valid
when used inside DNS. The proposals will also be very heavy tailed. When
DNS is exploring a distribution close to the prior, large proposals will
generally be needed. When DNS is exploring a distribution constrained to
very high values of the likelihood function, much smaller proposals will be
appropriate. Rather than attempting expensive tuning (since the number of
distributions involved may number in the hundreds or even thousands), we will
apply the heavy tailed proposal distributions and simply accept that there will
be some waste.

@@

\subsection{Proposals that modify $N$}\label{sec:proposal1}
The first kind of proposal we consider are proposals that change the
dimension of the model, i.e. proposals that change the value of $N$. By
necessity, these will also change the parameters $\{\xx_i\}_{i=1}^N$ because
the number of parameters will be changed. The proposals used here are
traditionally called {\it birth and death} proposals.

@@

If $N' > N$, the extra parameters needed, $\{\xx_i\}_{i=N+1}^{N'}$,
are drawn from their prior conditional on the current value of the
hyperparameters, i.e. the interim prior $p(x | \hyperparams)$.

@@

\subsection{Proposals that modify objects}\label{sec:proposal2}
We now consider a proposal distribution that modifies one or more of the
objects $\{\xx_i\}$, while keeping the number of objects $N$, as well as the
hyperparameters $\hyperparams$, fixed.

Let $F(x; \hyperparams)$ be a function that takes an object $x$ and transforms it
to a value $u$ that has a uniform distribution between 0 and 1, given $\hyperparams$.
If the object $x$ consists of a single scalar value, $F$ is the cumulative
distribution of the interim prior. Denote the inverse of $F$ by $G$.

A proposal
for an object involves transforming its parameters to $[0, 1]$ using $F$,
making the proposal in terms of $u$, and then transforming back.
Specifically, the proposal chooses
a new value $\xx_i'$ from the current value $\xx_i$ as follows:
\begin{eqnarray}
u_i &:=& F(\xx_i; \hyperparams)\\
u_i' &:=& \textnormal{mod}\left(u_i + \delta_u, 1\right)\\
\xx_i' &:=& G(u_i'; \hyperparams).
\end{eqnarray}
where $\delta_u$ is drawn from a heavy-tailed distribution.

@@

\subsection{Proposals that change the hyperparameters,
keeping the objects fixed}\label{sec:proposal3}
Another kind of proposal that we will consider is a proposal that keeps all of
the objects fixed in place (i.e. leaves $\{\xx_i\}$) unchanged, but changes
the hyperparameter(s) from their current value $\hyperparams$
to a new value $\hyperparams'$. The proposal for the hyperparameter(s) is chosen
to satisfy detailed balance with respect to $p(\hyperparams)$.

The overall Metropolis acceptance probability
for this kind of move, if sampling the prior (or the constrained prior of DNS)
must include the following factor:
\begin{eqnarray}
\frac{\prod_{i=1}^N p(\xx_i | \hyperparams')}{\prod_{i=1}^N p(\xx_i | \hyperparams)}
\label{eqn:acceptance_prob}
\end{eqnarray}
Since this proposal leaves the objects $\{\xx_i\}$ fixed, it will usually not
affect the value of the likelihood, and therefore the likelihood will not need
to be recomputed.

\subsection{Proposals that change the hyperparameters
and all of the objects}\label{sec:proposal4}
The above proposals allow for changes to $N$, $\hyperparams$, and $\{\xx_i\}$,
and are therefore sufficient to allow for ``correct'' exploration of the
prior distribution, and the constrained prior distributions used in DNS.
However, they do not necessarily allow for {\it efficient} exploration, even
of the prior itself. The main reason for this is the inability for large
changes to be made to the hyperparameters $\hyperparams$. If the proposed change
from $\hyperparams$ to $\hyperparams'$ is large, the ratio in
Equation~\ref{eqn:acceptance_prob} is likely to be very small, and the move
will probably be rejected.

Therefore, we allow an additional move that changes $\hyperparams$ to a new
value $\hyperparams'$, but rather than leaving the objects $\{\xx_i\}$ fixed, we
drag them so they represent the distribution $p(x | \hyperparams')$ rather than
$p(x|\hyperparams)$. We do this by making use of the transformation functions
$F(x; \hyperparams)$ and $G(x; \hyperparams)$ defined in Section~\ref{sec:proposal2}.

The ``dragging'' process works as follows, and must be carried out on
each object:
\begin{eqnarray}
u_i &:=& F(\xx_i; \hyperparams)\\
\xx_i' &:=& G(\xx_i; \hyperparams')
\end{eqnarray}
